Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
Update WSSSPE1_summary_report.tex
Added subsubsection on Licensing and Patents (expanded on what was there 
under "Industry").

Added short statement from my paper on how multi- and many core
computing forces the need for awareness of locality and synch.

Added short statement on licensing and patents and how they impact funding for SW

Communities are extremely important in software projects, and both
their building and continued engagement need attention during the
project life cycle.  Several of the submitted papers \todo{cite}
discussed how communities have been built around projects, and what is
needed to enable a project to grow. The latter includes public
source-code hosting, mailing lists, documentation, wikis, bug
trackers, software downloads, continuous integration, software quality
dashboards, and of course, a general web presence to tie a project's
channels and artifacts together.

@@

The papers submitted to this panel included many general
recommendations for processes, practices, tools, etc. One of the
papers \note{cite?}  suggested that a ``Software Sustainability
Institute'' should be vested with the authority to develop
standardized quality processes, a common repository, central resources
for services and consulting, a think tank of sorts, and a software
orphanage center (i.e., a place to `take care' of software when the
original developers have stopped doing so).  The idea of one common
repository received some resistance, with so many compelling
alternatives available, e.g., Bitbucket or GitHub. The centralized
communication or point of contact was seen as reasonable, with the
statement that ``vested with authority'' is perhaps too strong, but
``providing tools if needed'' might be more appropriate.

@@

These papers, and the discussion at the workshop, join a nascent
literature seeking to understand what drives software work in science
and how the reward systems of science thereby shape the type of
software work undertaken. This study includes the extent to which developers are
motivated to build software for the long-term, for the use of others,
and whether to work collaboratively or
separately~\cite{howison_incentives_2013, howison_scientific_2011,
  bietz_synergizing_2010}. Software work is not only motivated by
direct citations, but the visibility of software work in the
literature is important to those who write software used in science.

@@

While many papers briefly discussed licensing issues,
Elster~\cite{Elster_WSSSPE} directly discussed the impact of different
licenses on obtaining industrial funding for scientific software
projects.  In particular, she describes experiences with researchers
unwilling to utilize GPL (copyleft) code, as it adds restrictions to
reuse that some industries find unacceptable for future
commercialization.  (This was discussed by Hanwell et
al.~\cite{Hanwell_WSSSPE} as well.)  Further obstacles to
sustainability identified by the author include patenting of software,
a lack of open access, and even more broadly, a lack of open source
codes even in open access journals.  Finally, a lack of awareness on
the part of scientific software developers of commodity libraries for
common tasks reduces their ability to reuse code.

@@

@@

To conclude, we discuss what we have learned from the workshop and what we plan to do going forward.

@@

\begin{enumerate}
\item Use of shared repositories in the development of collaborative
  projects facilitates collaboration, reproducibility, and
  sustainability in computational science. However it represents a
  barrier in some scientific fields and has yet to be more widely
  adopted.

@@

% Conclusion of S8 : Use Cases, not sure where to put
%\note{conclusions from S8, not sure of location --Ketan}

@@

@@

\item Anne C. Elster. Software for Science: Some Personal
  Reflections~\cite{Elster_WSSSPE}

>>>>

Communities are extremely important in software projects, and both
their building and continued engagement need attention during the
project life cycle.  Several of the submitted papers 
% need help here...
\todo{cite}
discussed how communities have been built around projects, and what is
needed to enable a project to grow. The latter includes public
source-code hosting, mailing lists, documentation, wikis, bug
trackers, software downloads, continuous integration, software quality
dashboards, and of course, a general web presence to tie a project's
channels and artifacts together.

@@

The papers submitted to this panel included many general
recommendations for processes, practices, tools, etc. One of the
papers 
%N a todo!
\note{cite?}  
suggested that a ``Software Sustainability
Institute'' should be vested with the authority to develop
standardized quality processes, a common repository, central resources
for services and consulting, a think tank of sorts, and a software
orphanage center (i.e., a place to `take care' of software when the
original developers have stopped doing so).  The idea of one common
repository received some resistance, with so many compelling
alternatives available, e.g., Bitbucket or GitHub. The centralized
communication or point of contact was seen as reasonable, with the
statement that ``vested with authority'' is perhaps too strong. However,
``providing tools if needed'' might be more appropriate.

@@

%Anne: The folowing sentence is verbatim from the original Abstract..
% Se preivious commit for suggestion on hwo to shorten this sentence
These papers, and the discussion at the workshop, join a nascent
literature seeking to understand what drives software work in science
and how the reward systems of science thereby shape the type of
software work undertaken. This study includes the extent to which developers are
motivated to build software for the long-term, for the use of others,
and whether to work collaboratively or
separately~\cite{howison_incentives_2013, howison_scientific_2011,
  bietz_synergizing_2010}. Software work is not only motivated by
direct citations, but the visibility of software work in the
literature is important to those who write software used in science.

@@

\subsubsection{Licensing and Software Patents}
Software is no longer either open or closed (only binaries available),
but also licensed and patented, which clearly also impact 
software sustainability.
While many papers briefly discussed licensing issues,
Elster~\cite{Elster_WSSSPE} directly discussed the impact of software licenses 
on obtaining industrial funding for scientific software
projects.  In particular, she describes her experiences with researchers
unwilling to utilize GPL (copyleft) code, as it adds restrictions to
reuse that they themselve as we well as some industries find unacceptable for future
commercialization.  (This was discussed by Hanwell et
al.~\cite{Hanwell_WSSSPE} as well.)  US ICT companies funding 
academics will thus often insist on BSD licensing on software 
since they then can legally include the code into their commercial
codes.

On the other hand, there are companies
that fund larger GPL-licensed software projects \cite{Blatt_WSSSPE}
that insists that the university projects they fund 
also produce code with GPL licensing.
They do not accept BSD-like licenses since they argue that
other companies then may choose to build closed commercial
codes on they software they funded, rather than encouraging 
the community to contribute freely and thus ensuring software 
sustainability for the community.

In either case, the university researchers are not given much choice
if they want these much sought after funds in a increasingly competitive
grant world.

Another obstacles to
sustainability identified by the author include patenting of software.
Most countries place some limits on software patents. The European 
Union outright forbids them, while US patent law excludes “abstract ideas” 
that has been used to refuse some software patents.

Further obstacles to
sustainability is a lack of open access, and even more broadly, a lack of open source
codes even in open access journals.  Finally, a lack of awareness on
the part of scientific software developers of commodity libraries for
common tasks reduces their ability to reuse code.

@@

Elster \cite{Elster_WSSSPE} also points out how the 
prevalence and rapid growth of multi and many-core systems forces 
awareness of data locality and synchronization issues
if one want to teach people how to develop high-performing scientific codes.

@@

To conclude, we highlight what we have learned from the workshop, and what we plan to do going forward.

@@

\begin{enumerate}
\item Use of shared repositories in the development of collaborative
  projects facilitates collaboration, reproducibility, and
  sustainability in computational science. However, it represents a
  barrier in some scientific fields and has yet to be more widely
  adopted.

@@

\end{enumerate}
In addition, licensing and patents and how they impact research funding
for software developement, was also discussed.
% Conclusion of S8 : Use Cases, not sure where to put
%\note{conclusions from S8, not sure of location --Ketan}

@@

Elster would like to thank Statoil ASA for supporting her travel
and her research group through a grant related to the OPM/DUNE open source projects.

@@

\item Anne C. Elster. Software for Science: Some Personal
  Reflections on Funding, Licensing, Publishing and Teaching~\cite{Elster_WSSSPE}
